Implicit neural representation of potential fields

### 2024-06-06 intro.tex
Final read-through changes

\section{Magnetic potential fields}
The Earth has a naturally occuring magnetic field, approximating a dipole oriented toward the top and bottom of the planet.
Several properties of the magnetic field are important to the machine learning methods in this thesis.
First, that a field refers to a set of functions of space and time.
The magnetic field is a vector field, with three orthogonal components.
Neural networks with sufficient parameters are recognised as universal function approximators \parencite{hornikMultilayerFeedforwardNetworks1989}, and should therefore be capable of approximating the functions of the magnetic field.
Second, the curl of the magnetic vector field is a force function known as the magnetic potential field \parencite{kelloggFoundationsPotentialTheory1967}.
The magnitude of the magnetic vector field is the total magnetic intensity (TMI) \parencite{blakelyPotentialTheoryGravity1996}, with the unit nanotesla (nT).
Magnetic surveys in exploration geophysics generally create TMI products, where TMI may be measured directly or calculated from measurements of each vector component.
To support various scientific endeavours, a model of this field is periodically constructed --- the International Geomagnetic Reference Field (IGRF).
The residual between the IGRF and measured TMI at a specific time and location is termed magnetic anomaly.
It may be negative, and can exceed several thousand nanotesla in the Earth's environment.
One of the key petrophysical properties influencing the magnetic anomaly is magnetic susceptibility, itself predominantly influenced by magnetite, a mineral commonly present across geological settings \parencite{clarkNotesRockMagnetization1991}.
Rocks may also be intrinsicly magnetised with remnant magnetisation.
Because of these known petrophysical variations in rock magnetisation, surveys of the magnetic field can delineate and differentiate stratigraphic units and infer their composition.

\section{Aeromagnetic surveys}
\label{sec:introgeo}
Airborne surveys of the Earth's magnetic or gravitational potential fields are routinely acquired at a range of scales at different stages of exploration, from pre-competitive regional data to stimulate exploration \parencite{howardAirborneGeophysicalCoverage2004}, through to targeted investigations by the exploration industry at the prospect scale of up to several square kilometres.
Surveys may also be collected on the ground, or less commonly in exploration by ship or satellite, with most processing methods and interpretation shared between all forms of acquisition.
Aeromagnetic surveys are among the lowest cost geophysical method \parencite{dentithGeophysicsMineralExploration2014}.
There are many aspects of survey design controlling the resolvable detail within survey data.
When acquiring data, aircraft fly straight transect lines at a safe elevation above topography, known as the topography drape.
The power of higher frequencies in potential fields falls off with increasing source-sensor distance.
As such the lower the survey elevation, the more high-frequency components present in the potential field will be captured by the sensor.
Surveys are flown with a heading in which data will be acquired along line.
For regional and contemporary surveys this is recommended to be North-South, however, early practice was to survey perpendicular to the dominant geological strike, under the expectation of sampling the greatest frequency of change in magnetic intensity \parencite{islesGeologicalInterpretationAeromagnetic2013}.
Acquisition along flight lines is sampled at a rate of \qty{10}{\hertz} or greater, which corresponds to an interval of \qty{7}{\m} or shorter at nominal flight speeds of \qty{70}{\m\per\s} \parencite{goodwinAirborneMagneticRadiometric2023}.
The spacing between lines is a factor of the scale of investigation and the cost of acquisition, and regional pre-competitive data in Australia targets state-wide coverage at \qty{400}{\m}, which is generally regarded as the upper limit of usefulness to exploration \parencite{howardAirborneGeophysicalCoverage2004}.
These low-resolution surveys, funded by government geoscience agencies, are a key driver in greenfields exploration and are released as open file data.
Survey cost rapidly increases when surveying at higher resolution, due to overheads in flight line sampling.
Higher-resolution surveys over prospective areas fly lines at closer line spacings, of the order \qty{100}{\m}, in order to resolve greater detail.
The importance of these parameters in relation to resolvable detail will be outlined shortly.
The collected data may contain millions of samples scattered in three dimensions, and interpretation or processing often relies on regularisation to a quantised grid.

>>>>

\section{Magnetic potential fields}
The Earth has a naturally occuring magnetic field, approximating a dipole oriented toward the top and bottom of the planet.
Several properties of the magnetic field are important to the machine learning methods in this thesis.
First, that a field refers to a set of functions of space and time.
The magnetic field is a vector field, with three orthogonal components.
Neural networks with sufficient parameters are recognised as universal function approximators \parencite{hornikMultilayerFeedforwardNetworks1989}, and should therefore be capable of approximating the functions of the magnetic field.
Second, the curl of the magnetic vector field is a force function known as the magnetic potential field \parencite{kelloggFoundationsPotentialTheory1967}.
The magnitude of the magnetic vector field is the total magnetic intensity (TMI) \parencite{blakelyPotentialTheoryGravity1996}, with the unit nanotesla (nT).

Magnetic surveys in exploration geophysics generally create TMI products, where TMI may be measured directly or calculated from measurements of each vector component.
To support various scientific endeavours, a model of this field is periodically constructed --- the International Geomagnetic Reference Field (IGRF).
The residual between the IGRF and measured TMI at a specific time and location is termed magnetic anomaly.
It may be negative, and can exceed several thousand nanotesla in the Earth's environment.

One of the key petrophysical properties influencing the magnetic anomaly is magnetic susceptibility, itself predominantly influenced by magnetite, a mineral commonly present across geological settings \parencite{clarkNotesRockMagnetization1991}.
Rocks may also be intrinsicly magnetised with remnant magnetisation.
Because of these known petrophysical variations in rock magnetisation, surveys of the magnetic field can delineate and differentiate stratigraphic units and infer their composition.

\section{Aeromagnetic surveys}
\label{sec:introgeo}
Airborne surveys of the Earth's magnetic or gravitational potential fields are routinely acquired at a range of scales at different stages of exploration, from pre-competitive regional data to stimulate exploration \parencite{howardAirborneGeophysicalCoverage2004}, through to targeted investigations by the exploration industry at the prospect scale of up to several square kilometres.
Surveys may also be collected on the ground, or less commonly in exploration by ship or satellite, with most processing methods and interpretation shared between all forms of acquisition.
Aeromagnetic surveys are among the lowest cost geophysical method \parencite{dentithGeophysicsMineralExploration2014}.
Low-resolution regional surveys funded by government geoscience agencies are a key driver in greenfields exploration and are released as open file data.

There are many aspects of survey design controlling the resolvable detail within survey data.
When acquiring data, aircraft fly straight transect lines at a safe elevation above topography, known as the topography drape.
The power of higher frequencies in potential fields falls off with increasing source-sensor distance.
As such the lower the survey elevation, the more high-frequency components present in the potential field will be captured by the sensor.
Surveys are flown with a heading in which data will be acquired along line.
For regional and contemporary surveys this is recommended to be North-South, however, early practice was to survey perpendicular to the dominant geological strike, under the expectation of sampling the greatest frequency of change in magnetic intensity \parencite{islesGeologicalInterpretationAeromagnetic2013}.
Acquisition along flight lines is sampled at a rate of \qty{10}{\hertz} or greater, which corresponds to an interval of \qty{7}{\m} or shorter at nominal flight speeds of \qty{70}{\m\per\s} \parencite{goodwinAirborneMagneticRadiometric2023}.

The spacing between lines is a factor of the scale of investigation and the cost of acquisition, and regional pre-competitive data in Australia targets state-wide coverage at \qty{400}{\m}, which is generally regarded as the upper limit of usefulness to exploration \parencite{howardAirborneGeophysicalCoverage2004}.
Survey cost rapidly increases when surveying at higher resolution, due to overheads in flight line sampling.
Higher-resolution surveys over prospective areas fly lines at closer line spacings, of the order \qty{100}{\m}, in order to resolve greater detail.
The importance of these parameters in relation to resolvable detail will be outlined shortly.
The collected data may contain millions of samples scattered in three dimensions, and interpretation or processing often relies on regularisation to a quantised grid.

### 2024-06-06 paper1.tex
Final touch ups for typesetting and wording
Adjust chapter titles
Prompted by reviewer 2 to better present the thesis
Final read-through changes

\title{Magnetic grid resolution enhancement using machine learning: A case study from the Eastern Goldfields Superterrane}
\author[1*]{Luke Thomas Smith}
\author[1]{Tom Horrocks}
\author[1]{Eun-Jung Holden}
\author[1]{Daniel Wedge}
\author[2]{Naveed Akhtar}
\affil{Centre for Data-Driven Geoscience, School of Earth Sciences, The University of Western Australia (M006), 35 Stirling Highway, 6009 Perth, Australia}
\affil{Department of Computer Science and Software Engineering, The University of Western Australia (M002), 35 Stirling Highway, 6009 Perth, Australia}
\date{2022-09-21}
% \maketitle{}

@@

\begin{figure}[hbtp]
    \centering
    \includegraphics[width=0.75\linewidth]{fig/p1/convs_vec.pdf}
    \caption[RDB convolutional layers]{Each Residual Dense Block (\Cref{fig:rdb}, purple) comprises four convolutional layers, with additional learning pathways present as residual connections (black), and dense connections (green). LReLU refers to the Leaky-ReLU activation function. Adapted from \textcite{rakotonirinaESRGANFurtherImproving2020}.}
    \label{fig:convs}
\end{figure}

ESRGAN+, based on ESRGAN \parencite{wangESRGANEnhancedSuperresolution2018}, was chosen as the primary super-resolution CNN architecture for this task due to its reported success in other super-resolution tasks (especially \textcite{bodeUsingPhysicsInformedSuperResolution2019} and \textcite{leongDeepBedMapDeepNeural2020}) and state of the art results for its original application in image super-resolution.
As a Generative Adversarial Network (GAN), it comprises two iteratively trained neural networks, a Generator (G) used to upscale low-resolution inputs, and Discriminator (D), which is trained to improve the performance of the generator.

@@

\begin{table}[hbtp]
    \centering
    \begin{tblr}{
            colspec={
                    r |
                    Q[si={table-format=2.3},r]
                    Q[si={table-format=2.3},l] |
                    Q[si={table-format=2.3},r]
                    Q[si={table-format=2.3},l] |
                    Q[si={table-format=2.3},r]
                    Q[si={table-format=2.3},l]
                },
            row{1} = {guard}
        }
                & \SetCell[c=2]{c} Bicubic &            & \SetCell[c=2]{c} RDN\textdaggerdbl{} &                 & \SetCell[c=2]{c} ESRGAN+              \\
        \hline{}
                & \(\mu\)                  & \(\sigma\) & \(\mu\)                              & \(\sigma\)      & \(\mu\)                  & \(\sigma\) \\
        RMSE    & 13.545                   & 26.576     & \textbf{7.265}                       & \textbf{13.741} & 11.120                   & 21.670     \\
        SISM    & 0.924                    & 0.047      & \textbf{0.959}                       & \textbf{0.037}  & 0.922                    & 0.054      \\
        MS-SSIM & 0.971                    & 0.018      & \textbf{0.988}                       & \textbf{0.012}  & 0.976                    & 0.020      \\
    \end{tblr}

@@

\begin{landscape}
    \begin{figure}[hbtp]
        \centering
        \includegraphics[width=\linewidth,trim={0 0 0 -5mm},clip]{fig/p1/hpfilt.jpg}
        \caption[High-pass filtered results]{High-pass filtered test tile and its upsampled counterparts.
            The cut-off for the high-pass Butterworth filter was selected with wavelength equal to the Nyquist wavelength of the LR grid, i.e.,\ 4 times the HR Nyquist wavelength.
        }
        \label{fig:filtered}
    \end{figure}
\end{landscape}

>>>>

\title{Super-resolution for enhancing magnetic grids}
\author[1*]{Luke Thomas Smith}
\author[1]{Tom Horrocks}
\author[1]{Eun-Jung Holden}
\author[1]{Daniel Wedge}
\author[2]{Naveed Akhtar}
\affil{Centre for Data-Driven Geoscience, School of Earth Sciences, The University of Western Australia (M006), 35 Stirling Highway, 6009 Perth, Australia}
\affil{Department of Computer Science and Software Engineering, The University of Western Australia (M002), 35 Stirling Highway, 6009 Perth, Australia}
\date{2022-09-21}
% \maketitle{}

@@

\begin{figure}[hbtp]
    \centering
    \includegraphics[width=0.75\linewidth]{fig/p1/convs_vec.pdf}
    \caption[RDB convolutional layers]{Each Residual Dense Block (\Cref{fig:rdb}, purple) comprises four convolutional layers, with additional learning pathways present as residual connections (black), and dense connections (green). LReLU refers to the Leaky-ReLU activation. Adapted from \textcite{rakotonirinaESRGANFurtherImproving2020}.}
    \label{fig:convs}
\end{figure}

ESRGAN+, based on ESRGAN \parencite{wangESRGANEnhancedSuperresolution2018}, was chosen as the primary super-resolution CNN architecture for this task due to its reported success in other super-resolution tasks (especially \textcite{bodeUsingPhysicsinformedEnhanced2021} and \textcite{leongDeepBedMapDeepNeural2020}) and state of the art results for its original application in image super-resolution.
As a Generative Adversarial Network (GAN), it comprises two iteratively trained neural networks, a Generator (G) used to upscale low-resolution inputs, and Discriminator (D), which is trained to improve the performance of the generator.

@@

% \begin{landscape}
    \begin{figure}[hbtp]
        \centering
        \includegraphics[width=\linewidth,trim={0 0 0 -5mm},clip]{fig/p1/hpfilt.jpg}
        \caption[High-pass filtered results]{High-pass filtered test tile and its upsampled counterparts.
            The cut-off for the high-pass Butterworth filter was selected with wavelength equal to the Nyquist wavelength of the LR grid, i.e.,\ 4 times the HR Nyquist wavelength.
        }
        \label{fig:filtered}
    \end{figure}
% \end{landscape}

\begin{table}[hbtp]
    \centering
    \begin{tblr}{
            colspec={
                    r |
                    Q[si={table-format=2.3},r]
                    Q[si={table-format=2.3},l] |
                    Q[si={table-format=2.3},r]
                    Q[si={table-format=2.3},l] |
                    Q[si={table-format=2.3},r]
                    Q[si={table-format=2.3},l]
                },
            row{1} = {guard}
        }
                & \SetCell[c=2]{c} Bicubic &            & \SetCell[c=2]{c} RDN\textdaggerdbl{} &                 & \SetCell[c=2]{c} ESRGAN+              \\
        \hline{}
                & \(\mu\)                  & \(\sigma\) & \(\mu\)                              & \(\sigma\)      & \(\mu\)                  & \(\sigma\) \\
        RMSE    & 13.545                   & 26.576     & \textbf{7.265}                       & \textbf{13.741} & 11.120                   & 21.670     \\
        SSIM    & 0.924                    & 0.047      & \textbf{0.959}                       & \textbf{0.037}  & 0.922                    & 0.054      \\
        MS-SSIM & 0.971                    & 0.018      & \textbf{0.988}                       & \textbf{0.012}  & 0.976                    & 0.020      \\
    \end{tblr}

@@

### 2024-06-06 paper3.tex
Mention vertical gradient follow-on
Final read-through changes

Both the line data and grid raster netCDFs for the Wolfe Creek survey were downloaded from Geoscience Australia (GA) \footnote{https://pid.geoscience.gov.au/dataset/ga/142694} and used without further processing.
The line data netCDF contains tie-levelled, micro-levelled, and AWAGS levelled magnetic variables, in increasing stages of processing respectively.
The micro-levelled magnetic data are used in this case study because the originally recorded sample altitude is uncertain in the available levelled data.
Either a geographic and projected coordinate system can be used, provided for consistency between the query coordinates for inference and those used during training.
The three spatial coordinate dimensions are independently and linearly min-max normalised between \num{-1} and \num{1}.
While this means the axes in the latent spatial domain may be compressed relative to each other axis, the inverse normalisation process returns proportionality between axes.
The case study data do not span a large or magnetically varying geological extent, hence it is sufficient to use simple min-max normalisation for the magnetic field data within the global extent of the survey samples, here between \num{-1} and \num{1}.
If extensive or complexly varying magnetic provinces were present within the dataset, a normalisation method more robust to outliers may be required.
The \num{76420} sample points contained within the case study line data are split, with \qty{70}{\percent} used for training, and a random \qty{20}{\percent} of the remaining points are reserved to validate the model performance during training.
Inference is performed by querying novel coordinates within the learned domain, here at regularly spaced grid intervals for comparison with standard grid regularisation methods.

@@

\section{Discussion}
\label{sec:3discussion}
The proposed method has strong learning capacity for the potential field samples in the horizontal plane near the nominal survey altitude, allowing extraction of level grids which are accurate against the reference grid.
Level grids can be queried at arbitrary heights and cell sizes, with the highest accuracy achieved at the realised nominal survey altitude.
Performance decreases within approximately one standard deviation from that height (\qty{35} to \qty{45}{\m}), with gradual failure of the model outside of that range.
It is interpreted from the poor quality of the vertical gradients and out-of-range level grid predictions that the model is learning a function between horizontally coplanar points, and not an accurate vertical continuation function.
This is conjectured to be caused by aeromagnetic sampling being carried out on a nominally horizontal planar drape surface, without sufficient repeat measurements at other heights to train the network for the vertical field component.

>>>>

Both the line data and grid raster netCDFs for the Wolfe Creek survey were downloaded from Geoscience Australia (GA)\footnote{https://pid.geoscience.gov.au/dataset/ga/142694} and used without further processing.
The line data netCDF contains tie-levelled, micro-levelled, and AWAGS levelled magnetic variables, in increasing stages of processing respectively.
The micro-levelled magnetic data are used in this case study because the originally recorded sample altitude is uncertain in the available levelled data.
Either a geographic or projected coordinate system can be used, provided for consistency between the query coordinates for inference and those used during training.
The three spatial coordinate dimensions are independently and linearly min-max normalised between \num{-1} and \num{1}.
While this means the axes in the latent spatial domain may be compressed relative to each other axis, the inverse normalisation process returns proportionality between axes.
The case study data do not span a large or magnetically varying geological extent, hence it is sufficient to use simple min-max normalisation for the magnetic field data within the global extent of the survey samples, here between \num{-1} and \num{1}.
If extensive or complexly varying magnetic provinces were present within the dataset, a normalisation method more robust to outliers may be required.
The \num{76420} sample points contained within the case study line data are split, with \qty{70}{\percent} used for training, and a random \qty{20}{\percent} of the remaining points are reserved to validate the model performance during training.
Inference is performed by querying novel coordinates within the learned domain, here at regularly spaced grid intervals for comparison with standard grid regularisation methods.

@@

\section{Discussion}
\label{sec:3discussion}
The proposed method has strong learning capacity for the potential field samples in the horizontal plane near the nominal survey altitude, allowing extraction of level grids which are accurate against the reference grid.
Level grids can be queried at arbitrary heights and cell sizes, with the highest accuracy achieved at the realised nominal survey altitude.
Performance decreases within approximately one standard deviation from that height (\qty{35} to \qty{45}{\m}), with gradual failure of the model outside of that range.
It is interpreted from the poor quality of the vertical gradients and out-of-range level grid predictions that the model is learning a function between horizontally coplanar points, and not an accurate vertical continuation function.
This is conjectured to be caused by aeromagnetic sampling being carried out on a nominally horizontal planar drape surface, without sufficient repeat measurements at other heights to train the network for the vertical field component.
Initial experiments with synthetic data evenly sampled across the vertical extent learn a vertical gradient with similar performance to the horizontal gradients, supporting this conjecture.
